---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data

Define global options for the whole document and required libraries are loaded.
Data are read from the zipped file.
```{r global_options}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, output=TRUE)
```

```{r}
library(data.table)
library(ggplot2)
library(chron)
library(plyr)
library(gridExtra)


unzip('./activity.zip')
file <- './activity.csv'

data <- data.table(read.csv(file))
data$date <- as.Date(data$date)
```

## What is mean total number of steps taken per day?

Histogram of number of steps per day
```{r}
daySteps <- data[,list(total=sum(steps)),by=date]
plot1 <- qplot(daySteps$total,binwidth=10) + 
        geom_histogram(binwidth=500) +
        xlab("# steps per day") +
        ylab("count")
plot1
```

Mean and median number of steps per day:
```{r}
mean(daySteps$total)
median(daySteps$total)
```

## What is the average daily activity pattern?

Time series for the average number of steps during the whole day, represented by the 5-min time interval
```{r}
dayActivity <- data[,list(total=mean(steps,na.rm=TRUE)),by=interval]
qplot(interval, total, data=dayActivity, geom="line")+
        xlab("5 min day interval") +
        ylab("Average number of steps")
```

Maximum number of steps reached on average at 5-minute interval
```{r}
dayActivity$interval[dayActivity$total == max(dayActivity$total)]
```


## Imputing missing values

Detect the total number of missing values in the dataset
```{r}
idx <- is.na(data$steps)
sum(idx)
```

Filling in all of the missing values in the dataset. The average value for the interval is employed.
```{r}
# Show some of the missing step values before filling in
data[idx,]

# Get corresponding interval index and fill in the missing values with the average number of steps
data$steps[idx] <- dayActivity$total[match(data$interval[idx],dayActivity$interval)]

# Show some of the missing step values after filling in
data[idx,]
```

Histogram of number of steps per day after filling in NA's (comparison against previous one,left)
```{r}
daySteps2 <- data[,list(total=sum(steps)),by=date]
plot2 <- qplot(daySteps2$total,binwidth=10) + 
        geom_histogram(binwidth=500) +
        xlab("# steps per day") +
        ylab("count") +
        ggtitle("Filled NA's")

plot1 <- plot1 + ggtitle("NA's present")

grid.arrange(plot1, plot2, ncol=2)
```

Mean and median number of steps per day after filling in NA's
```{r}
mean(daySteps2$total)
median(daySteps2$total)
```

## Are there differences in activity patterns between weekdays and weekends?

In order to observe differences, the average activity pattern is computed segregated for weedays and weekends
```{r}
# Detect wether a day is week or weekend day
data$dayType <- as.factor(is.weekend(data$date))
levels(data$dayType)[levels(data$dayType)==TRUE]  <- "Weekend"
levels(data$dayType)[levels(data$dayType)==FALSE] <- "Weekday"

# Segment average activity by interval and day type
dayActivity2 <- data[,list(total=mean(steps,na.rm=TRUE)),by=list(interval,dayType)]

# Ploting
qplot(interval, total, data=dayActivity2, geom="line") + facet_grid(dayType~.) +
        xlab("5 min day interval") +
        ylab("Average number of steps")
```


